---
title: "QC v9: Pearson-correlation technical-replicate filtering"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
---

# Purpose

This is step 2 of the v9 pipeline. It uses the cleaned evidence and experimental design already prepared by `1.Basic_Info_v9.Rmd`. For each biological experiment, a median pairwise Pearson correlation below 0.7 triggers removal of the technical replicate with the lowest median correlation to the other technical replicates.

# Settings

```{r settings}
load_project_packages(c("data.table", "ggplot2"))

root <- basic_obj$project_root
pth <- list(
  qc_dir = file.path(root, "QC_v9"),
  qc_table = file.path(root, "QC_v9", "tables"),
  qc_plots = file.path(root, "QC_v9", "plots")
)

dir.create(pth$qc_table, recursive = TRUE, showWarnings = FALSE)
dir.create(pth$qc_plots, recursive = TRUE, showWarnings = FALSE)

raw0 <- data.table::copy(basic_obj$raw_map)
raw0[SourceBlock == "20260331", SourceBlock := "RS"]
cond <- as.character(basic_obj$condition_order)
pearson_threshold <- 0.7

condition_colors <- stats::setNames(
  c("#0072B2", "#D55E00", "#009E73", "#CC79A7", "#E69F00", "#56B4E9", "#F0E442", "#000000")[seq_along(cond)],
  cond
)
```

# QC functions

```{r qc_functions}
make_protein_rawfile_table <- function(clean_evidence, raw_map) {
  protein_table <- clean_evidence[
    RawFile %in% raw_map$RawFile &
      !is.na(LeadingProteinForCounting) &
      LeadingProteinForCounting != "",
    .(ProteinIntensity = sum(Intensity, na.rm = TRUE)),
    by = .(RawFile, ProteinGroup = LeadingProteinForCounting)
  ]

  merge(
    protein_table,
    raw_map[, .(RawFile, SourceBlock, Condition, BioRep, BioExperiment, TechRep)],
    by = "RawFile",
    sort = FALSE
  )
}

make_pairwise_tech_cor <- function(protein_table, raw_map) {
  data.table::rbindlist(lapply(unique(raw_map$BioExperiment), function(bio_id) {
    meta <- raw_map[BioExperiment == bio_id][order(TechRep)]
    wide <- data.table::dcast(
      protein_table[RawFile %in% meta$RawFile],
      ProteinGroup ~ RawFile,
      value.var = "ProteinIntensity",
      fill = NA_real_
    )

    data.table::rbindlist(lapply(utils::combn(meta$RawFile, 2, simplify = FALSE), function(pair) {
      x <- wide[[pair[1]]]
      y <- wide[[pair[2]]]
      keep <- stats::complete.cases(x, y) & x > 0 & y > 0
      m1 <- meta[RawFile == pair[1]]
      m2 <- meta[RawFile == pair[2]]

      data.table::data.table(
        Groups = as.character(m1$SourceBlock),
        Condition = as.character(m1$Condition),
        BioExperiment = as.character(bio_id),
        BioRep = as.character(m1$BioRep),
        RawFile_1 = pair[1],
        RawFile_2 = pair[2],
        TechRep_1 = as.character(m1$TechRep),
        TechRep_2 = as.character(m2$TechRep),
        Pearson_r = stats::cor(log2(x[keep]), log2(y[keep]), method = "pearson")
      )
    }))
  }))
}

make_total_intensity <- function(protein_table, raw_map) {
  totals <- protein_table[, .(
    Total_protein_intensity = sum(ProteinIntensity, na.rm = TRUE)
  ), by = RawFile]

  merge(
    totals,
    raw_map[, .(RawFile, Groups = SourceBlock, Condition, BioExperiment, TechRep)],
    by = "RawFile",
    sort = FALSE
  )
}

save_qc_plot <- function(plot, name, width = 12, height = 8) {
  ggplot2::ggsave(
    file.path(pth$qc_plots, paste0(name, ".png")),
    plot,
    width = width,
    height = height,
    dpi = 300
  )
}
```

# Pearson-correlation filtering

```{r pearson_filtering}
prot_raw0 <- make_protein_rawfile_table(clean_evi, raw0)
pair_cor_before <- make_pairwise_tech_cor(prot_raw0, raw0)

bio_cor_before <- pair_cor_before[, .(
  Experiment_median_Pearson = stats::median(Pearson_r, na.rm = TRUE)
), by = .(Groups, Condition, BioExperiment)]

rep_cor <- data.table::rbindlist(list(
  pair_cor_before[, .(
    Groups, Condition, BioExperiment,
    RawFile = RawFile_1,
    TechRep = TechRep_1,
    Pearson_r
  )],
  pair_cor_before[, .(
    Groups, Condition, BioExperiment,
    RawFile = RawFile_2,
    TechRep = TechRep_2,
    Pearson_r
  )]
))[, .(
  Replicate_median_Pearson = stats::median(Pearson_r, na.rm = TRUE)
), by = .(Groups, Condition, BioExperiment, RawFile, TechRep)]

tech_out <- merge(
  rep_cor,
  bio_cor_before,
  by = c("Groups", "Condition", "BioExperiment")
)

rm_tech <- tech_out[
  Experiment_median_Pearson < pearson_threshold
][order(BioExperiment, Replicate_median_Pearson, TechRep), .SD[1], by = BioExperiment]

tech_out[, Removed := RawFile %in% rm_tech$RawFile]
tech_out[, Pearson_threshold := pearson_threshold]
data.table::setorder(tech_out, Groups, Condition, BioExperiment, TechRep)

raw <- raw0[!RawFile %in% rm_tech$RawFile]
attr(raw, "condition_order") <- cond

data.table::fwrite(
  tech_out,
  file.path(pth$qc_table, "technical_outlier_report.csv")
)

rm_tech[, .(
  RawFile,
  Groups,
  Condition,
  BioExperiment,
  TechRep,
  Experiment_median_Pearson,
  Replicate_median_Pearson
)]
```

# QC data after filtering

```{r qc_data_after_filtering}
prot_raw <- make_protein_rawfile_table(clean_evi, raw)
pair_cor <- make_pairwise_tech_cor(prot_raw, raw)
total_int <- make_total_intensity(prot_raw, raw)

bio_total <- total_int[, .(
  Total_protein_intensity = stats::median(Total_protein_intensity, na.rm = TRUE)
), by = .(Groups, Condition, BioExperiment)]

bio_cor <- pair_cor[, .(
  Median_Pearson = stats::median(Pearson_r, na.rm = TRUE)
), by = .(Groups, Condition, BioExperiment)]

bio_order <- unique(raw[order(SourceBlock, Condition, BioExperiment)]$BioExperiment)
total_int[, BioExperiment := factor(BioExperiment, levels = bio_order)]
pair_cor[, BioExperiment := factor(BioExperiment, levels = bio_order)]
```

# Unique peptide and protein boxplots

```{r unique_peptide_protein_boxplots}
bio_info <- unique(raw[, .(Groups = SourceBlock, Condition, BioExperiment)])
bio_info[, Condition := factor(as.character(Condition), levels = cond)]

unique_counts <- clean_evi[RawFile %in% raw$RawFile, .(
  Unique_peptide_sequences = data.table::uniqueN(Sequence[!is.na(Sequence) & Sequence != ""]),
  Unique_leading_protein_groups = data.table::uniqueN(LeadingProteinForCounting[
    !is.na(LeadingProteinForCounting) & LeadingProteinForCounting != ""
  ])
), by = BioExperiment]

unique_counts <- merge(unique_counts, bio_info, by = "BioExperiment", sort = FALSE)

p_unique_pep <- ggplot2::ggplot(
  unique_counts,
  ggplot2::aes(x = Condition, y = Unique_peptide_sequences, fill = Condition)
) +
  ggplot2::geom_boxplot(width = 0.65, outlier.shape = NA) +
  ggplot2::geom_jitter(width = 0.10, size = 2.2, alpha = 0.85) +
  ggplot2::scale_fill_manual(values = condition_colors, drop = FALSE) +
  ggplot2::labs(
    title = "Unique peptide sequences by biological experiment",
    x = "Condition",
    y = "Unique peptide sequences",
    fill = "Condition"
  ) +
  ggplot2::theme_bw(base_size = 14) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 30, hjust = 1),
    legend.position = "none"
  )

p_unique_prot <- ggplot2::ggplot(
  unique_counts,
  ggplot2::aes(x = Condition, y = Unique_leading_protein_groups, fill = Condition)
) +
  ggplot2::geom_boxplot(width = 0.65, outlier.shape = NA) +
  ggplot2::geom_jitter(width = 0.10, size = 2.2, alpha = 0.85) +
  ggplot2::scale_fill_manual(values = condition_colors, drop = FALSE) +
  ggplot2::labs(
    title = "Unique leading protein groups by biological experiment",
    x = "Condition",
    y = "Unique leading protein groups",
    fill = "Condition"
  ) +
  ggplot2::theme_bw(base_size = 14) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 30, hjust = 1),
    legend.position = "none"
  )

save_qc_plot(p_unique_pep, "unique_peptide_sequences_by_condition")
save_qc_plot(p_unique_prot, "unique_leading_protein_groups_by_condition")

p_unique_pep
p_unique_prot
```

# Total intensity boxplots

```{r total_intensity_boxplots}
p_total_bio <- ggplot2::ggplot(
  total_int,
  ggplot2::aes(x = BioExperiment, y = Total_protein_intensity, fill = Condition)
) +
  ggplot2::geom_boxplot(width = 0.65, outlier.shape = NA) +
  ggplot2::geom_jitter(width = 0.10, size = 2, alpha = 0.85) +
  ggplot2::facet_grid(. ~ Groups, scales = "free_x", space = "free_x") +
  ggplot2::scale_fill_manual(values = condition_colors, drop = FALSE) +
  ggplot2::labs(
    title = "Total protein intensity by biological experiment",
    x = "Biological experiment",
    y = "Total protein intensity",
    fill = "Condition"
  ) +
  ggplot2::theme_bw(base_size = 14) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold", hjust = 0.5),
    axis.text.x = ggplot2::element_text(angle = 30, hjust = 1)
  )

p_total_condition <- ggplot2::ggplot(
  bio_total,
  ggplot2::aes(x = Condition, y = Total_protein_intensity, fill = Condition)
) +
  ggplot2::geom_boxplot(width = 0.65, outlier.shape = NA) +
  ggplot2::geom_jitter(width = 0.10, size = 2.2, alpha = 0.85) +
  ggplot2::scale_fill_manual(values = condition_colors, drop = FALSE) +
  ggplot2::labs(
    title = "Total protein intensity by condition",
    x = "Condition",
    y = "Median total protein intensity",
    fill = "Condition"
  ) +
  ggplot2::theme_bw(base_size = 14) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold", hjust = 0.5),
    axis.text.x = ggplot2::element_text(angle = 30, hjust = 1),
    legend.position = "none"
  )

save_qc_plot(p_total_bio, "total_intensity_by_bioexperiment", width = 16)
save_qc_plot(p_total_condition, "total_intensity_by_condition")

p_total_bio
p_total_condition
```

# Pearson correlation boxplots

```{r pearson_correlation_boxplots}
p_cor_bio <- ggplot2::ggplot(
  pair_cor,
  ggplot2::aes(x = BioExperiment, y = Pearson_r, fill = Condition)
) +
  ggplot2::geom_boxplot(width = 0.65, outlier.shape = NA) +
  ggplot2::geom_jitter(width = 0.10, size = 2, alpha = 0.85) +
  ggplot2::facet_grid(. ~ Groups, scales = "free_x", space = "free_x") +
  ggplot2::scale_fill_manual(values = condition_colors, drop = FALSE) +
  ggplot2::labs(
    title = "Pearson correlation by biological experiment",
    x = "Biological experiment",
    y = "Pearson correlation",
    fill = "Condition"
  ) +
  ggplot2::theme_bw(base_size = 14) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold", hjust = 0.5),
    axis.text.x = ggplot2::element_text(angle = 30, hjust = 1)
  )

p_cor_condition <- ggplot2::ggplot(
  bio_cor,
  ggplot2::aes(x = Condition, y = Median_Pearson, fill = Condition)
) +
  ggplot2::geom_boxplot(width = 0.65, outlier.shape = NA) +
  ggplot2::geom_jitter(width = 0.10, size = 2.2, alpha = 0.85) +
  ggplot2::scale_fill_manual(values = condition_colors, drop = FALSE) +
  ggplot2::labs(
    title = "Pearson correlation by condition",
    x = "Condition",
    y = "Median Pearson correlation",
    fill = "Condition"
  ) +
  ggplot2::theme_bw(base_size = 14) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold", hjust = 0.5),
    axis.text.x = ggplot2::element_text(angle = 30, hjust = 1),
    legend.position = "none"
  )

save_qc_plot(p_cor_bio, "pearson_correlation_by_bioexperiment", width = 16)
save_qc_plot(p_cor_condition, "pearson_correlation_by_condition")

p_cor_bio
p_cor_condition
```

# Before and after filtering within affected groups

```{r affected_group_before_after}
total_int0 <- make_total_intensity(prot_raw0, raw0)
affected_groups <- unique(as.character(rm_tech$Groups))

for (src in affected_groups) {
  total_before <- data.table::copy(total_int0[Groups == src])
  total_after <- data.table::copy(total_int[Groups == src])
  total_before[, Filter := "Before filtering"]
  total_after[, Filter := "After filtering"]
  total_plot <- data.table::rbindlist(list(total_before, total_after))
  total_plot[, Filter := factor(Filter, levels = c("Before filtering", "After filtering"))]

  p_total_filter <- ggplot2::ggplot(
    total_plot,
    ggplot2::aes(x = BioExperiment, y = Total_protein_intensity, fill = BioExperiment)
  ) +
    ggplot2::geom_boxplot(width = 0.65, outlier.shape = NA) +
    ggplot2::geom_jitter(width = 0.10, size = 2, alpha = 0.85) +
    ggplot2::facet_grid(. ~ Filter) +
    ggplot2::labs(
      title = paste0("Total protein intensity: ", src),
      x = "Biological experiment",
      y = "Total protein intensity",
      fill = "Biological experiment"
    ) +
    ggplot2::theme_bw(base_size = 14) +
    ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold", hjust = 0.5),
      axis.text.x = ggplot2::element_text(angle = 30, hjust = 1),
      legend.position = "bottom"
    )

  cor_before <- data.table::copy(pair_cor_before[Groups == src])
  cor_after <- data.table::copy(pair_cor[Groups == src])
  cor_before[, Filter := "Before filtering"]
  cor_after[, Filter := "After filtering"]
  cor_plot <- data.table::rbindlist(list(cor_before, cor_after), fill = TRUE)
  cor_plot[, Filter := factor(Filter, levels = c("Before filtering", "After filtering"))]

  p_cor_filter <- ggplot2::ggplot(
    cor_plot,
    ggplot2::aes(x = BioExperiment, y = Pearson_r, fill = BioExperiment)
  ) +
    ggplot2::geom_boxplot(width = 0.65, outlier.shape = NA) +
    ggplot2::geom_jitter(width = 0.10, size = 2, alpha = 0.85) +
    ggplot2::facet_grid(. ~ Filter) +
    ggplot2::labs(
      title = paste0("Pearson correlation: ", src),
      x = "Biological experiment",
      y = "Pearson correlation",
      fill = "Biological experiment"
    ) +
    ggplot2::theme_bw(base_size = 14) +
    ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold", hjust = 0.5),
      axis.text.x = ggplot2::element_text(angle = 30, hjust = 1),
      legend.position = "bottom"
    )

  group_name <- gsub("[^A-Za-z0-9_-]+", "_", src)
  save_qc_plot(p_total_filter, paste0("total_intensity_before_after_", group_name), width = 16)
  save_qc_plot(p_cor_filter, paste0("pearson_correlation_before_after_", group_name), width = 16)

  print(p_total_filter)
  print(p_cor_filter)
}
```

# Prepare QC object

```{r prepare_qc_object}
qc_obj <- list(
  raw_map_original = raw0,
  raw_map_filtered = raw,
  condition_order = cond,
  technical_outlier_report = tech_out,
  removed_technical_replicates = rm_tech,
  settings = list(
    workflow_version = "v9",
    filtering_rule = "experiment median Pearson < 0.7; remove replicate with lowest median Pearson",
    pearson_threshold = pearson_threshold
  )
)
```
