---
title: "Differential analysis v9: batch-corrected protein-level limma workflow"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
---

# Purpose

This is step 3 of the v9 pipeline. It uses the cleaned evidence from `1.Basic_Info_v9.Rmd` and the filtered technical-replicate design from `2.QC_v9.Rmd`, then performs peptide normalization, protein aggregation, group correction, and protein-level differential analysis.

# Settings

```{r settings}
load_project_packages(c(
  "data.table", "limma", "QFeatures", "SummarizedExperiment",
  "S4Vectors", "matrixStats"
))

root <- basic_obj$project_root
pth <- list(
  da_dir = file.path(root, "Differential_Analysis_v9"),
  da_table = file.path(root, "Differential_Analysis_v9", "tables"),
  da_plots = file.path(root, "Differential_Analysis_v9", "plots")
)

dir.create(pth$da_table, recursive = TRUE, showWarnings = FALSE)
dir.create(pth$da_plots, recursive = TRUE, showWarnings = FALSE)

raw0 <- data.table::copy(qc_obj$raw_map_original)
raw <- data.table::copy(qc_obj$raw_map_filtered)
cond <- as.character(qc_obj$condition_order)
tech_out <- data.table::copy(qc_obj$technical_outlier_report)
rm_tech <- data.table::copy(qc_obj$removed_technical_replicates)

raw0[SourceBlock == "20260331", SourceBlock := "RS"]
raw[SourceBlock == "20260331", SourceBlock := "RS"]
data.table::setnames(raw0, "SourceBlock", "Groups")
data.table::setnames(raw, "SourceBlock", "Groups")

attr(raw0, "condition_order") <- cond
attr(raw, "condition_order") <- cond

max_na <- 2 / 3
min_det <- 1
adj_m <- "BH"
alpha <- 0.05
lfc_thr <- 1
```

# Analysis functions

```{r analysis_functions}
first_value <- function(x) {
  x[!is.na(x) & x != ""][1]
}

collapse_values <- function(x) {
  paste(sort(unique(x[!is.na(x) & x != ""])), collapse = ";")
}

summarize_to_bio <- function(mat, raw_map) {
  bio_info <- unique(raw_map[, .(
    Groups,
    Condition,
    BioExperiment,
    BioRep
  )])
  data.table::setorder(bio_info, Condition, BioRep, BioExperiment)

  mat_bio <- vapply(
    bio_info$BioExperiment,
    function(bio_id) {
      matrixStats::rowMedians(
        mat[, raw_map[BioExperiment == bio_id, RawFile], drop = FALSE],
        na.rm = TRUE
      )
    },
    numeric(nrow(mat))
  )

  rownames(mat_bio) <- rownames(mat)
  colnames(mat_bio) <- bio_info$BioExperiment
  mat_bio[is.nan(mat_bio)] <- NA_real_

  list(mat = mat_bio, bio = bio_info)
}
```

# Comparisons

```{r comparisons}
cmp <- data.table::data.table(
  Target = c(
    "c-jun_KO",
    "CTRL_cDDP_06h", "CTRL_cDDP_24h",
    "c-jun_KO_cDDP_06h", "c-jun_KO_cDDP_24h",
    "c-jun_KO_cDDP_06h", "c-jun_KO_cDDP_24h",
    "CTRL_JNK-IN-8_cDDP_06h", "CTRL_JNK-IN-8_cDDP_24h",
    "CTRL_JNK-IN-8_cDDP_06h", "CTRL_JNK-IN-8_cDDP_24h",
    "CTRL_cDDP_24h", "c-jun_KO_cDDP_24h", "CTRL_JNK-IN-8_cDDP_24h"
  ),
  Reference = c(
    "CTRL",
    "CTRL", "CTRL",
    "c-jun_KO", "c-jun_KO",
    "CTRL_cDDP_06h", "CTRL_cDDP_24h",
    "CTRL_cDDP_06h", "CTRL_cDDP_24h",
    "c-jun_KO_cDDP_06h", "c-jun_KO_cDDP_24h",
    "CTRL_cDDP_06h", "c-jun_KO_cDDP_06h", "CTRL_JNK-IN-8_cDDP_06h"
  )
)

cmp[, ComparisonShortID := sprintf("C%02d", .I)]
cmp[, ComparisonID := paste0(Target, "_vs_", Reference)]
cmp[, Exploratory := .I >= 12]
```

# Peptide matrix

```{r peptide_matrix}
evi_ann <- data.table::copy(clean_evi[RawFile %in% raw$RawFile])

evi_ann[, ProteinID := data.table::fifelse(
  !is.na(`Leading proteins`) & `Leading proteins` != "",
  as.character(`Leading proteins`),
  as.character(Proteins)
)]
evi_ann[, PepSeq := as.character(Sequence)]
evi_ann <- evi_ann[
  !is.na(ProteinID) & ProteinID != "" &
    !is.na(PepSeq) & PepSeq != ""
]
evi_ann[, PepID := paste(ProteinID, PepSeq, sep = "|")]

pep_long <- evi_ann[, .(
  Intensity = sum(Intensity, na.rm = TRUE)
), by = .(PepID, RawFile)]

pep_wide <- data.table::dcast(
  pep_long,
  PepID ~ RawFile,
  value.var = "Intensity",
  fill = NA_real_
)

raw_keep <- raw[RawFile %in% names(pep_wide), RawFile]
pep_m <- as.matrix(pep_wide[, ..raw_keep])
rownames(pep_m) <- pep_wide$PepID
pep_m[pep_m <= 0] <- NA_real_
```

# Peptide annotation

```{r peptide_annotation}
pep_anno <- evi_ann[, .(
  Sequence = first_value(PepSeq),
  ProteinID = first_value(ProteinID),
  Proteins = collapse_values(Proteins),
  Gene.names = collapse_values(`Gene names`),
  Protein.names = collapse_values(`Protein names`)
), by = PepID]

pep_anno <- pep_anno[match(rownames(pep_m), PepID)]
raw_info <- raw[match(colnames(pep_m), RawFile)]

pep_se <- SummarizedExperiment::SummarizedExperiment(
  assays = list(intensity = pep_m),
  rowData = S4Vectors::DataFrame(
    as.data.frame(pep_anno[, .(
      Sequence, ProteinID, Proteins, Gene.names, Protein.names
    )])
  ),
  colData = S4Vectors::DataFrame(as.data.frame(raw_info))
)
rownames(pep_se) <- pep_anno$PepID
```

# Normalize and aggregate

```{r normalize_and_aggregate}
pep_f <- pep_se |>
  QFeatures::zeroIsNA() |>
  QFeatures::filterNA(pNA = max_na)

qf <- QFeatures::QFeatures(list(peptides = pep_f))
SummarizedExperiment::colData(qf) <- SummarizedExperiment::colData(qf[["peptides"]])

qf <- qf |>
  QFeatures::logTransform(i = "peptides", name = "log_peptides") |>
  QFeatures::normalize(
    i = "log_peptides",
    name = "lognorm_peptides",
    method = "center.median"
  ) |>
  QFeatures::aggregateFeatures(
    i = "lognorm_peptides",
    name = "proteins_raw",
    fcol = "ProteinID",
    fun = matrixStats::colMedians,
    na.rm = TRUE
  )

pep_raw <- SummarizedExperiment::assay(qf[["lognorm_peptides"]])
prot_raw <- SummarizedExperiment::assay(qf[["proteins_raw"]])

pep_bio <- summarize_to_bio(pep_raw, raw)
prot_bio <- summarize_to_bio(prot_raw, raw)

pep_bio_m <- pep_bio$mat
p_bio_m <- prot_bio$mat
bio <- prot_bio$bio
```

# Correct group effects

```{r group_correction}
condition_design <- stats::model.matrix(~ 0 + Condition, data = bio)

pep_bio_m <- limma::removeBatchEffect(
  pep_bio_m,
  batch = factor(bio$Groups),
  design = condition_design
)

p_bio_m <- limma::removeBatchEffect(
  p_bio_m,
  batch = factor(bio$Groups),
  design = condition_design
)
```

# Protein annotation

```{r protein_annotation}
pep_f_anno <- data.table::as.data.table(
  as.data.frame(SummarizedExperiment::rowData(qf[["lognorm_peptides"]]))
)

prot_anno <- pep_f_anno[
  !is.na(ProteinID) & ProteinID != "",
  .(
    Gene.names = collapse_values(Gene.names),
    Protein.names = collapse_values(Protein.names),
    Peptides_used = .N
  ),
  by = ProteinID
]
```

# Differential analysis

```{r differential_analysis}
condition_code <- stats::setNames(paste0("G", seq_along(cond)), cond)

run_comparison <- function(cmp_i) {
  target <- cmp_i$Target
  reference <- cmp_i$Reference
  ref_bio <- bio[as.character(Condition) == reference, BioExperiment]
  target_bio <- bio[as.character(Condition) == target, BioExperiment]
  sample_ids <- c(ref_bio, target_bio)

  mat0 <- p_bio_m[, sample_ids, drop = FALSE]
  n_ref <- rowSums(!is.na(mat0[, ref_bio, drop = FALSE]))
  n_target <- rowSums(!is.na(mat0[, target_bio, drop = FALSE]))
  keep <- n_ref >= min_det & n_target >= min_det
  mat <- mat0[keep, , drop = FALSE]

  model_data <- bio[match(sample_ids, BioExperiment)]
  model_data[, Code := condition_code[as.character(Condition)]]
  model_data[, ModelGroup := factor(
    Code,
    levels = condition_code[c(reference, target)]
  )]

  design <- stats::model.matrix(~ 0 + ModelGroup, data = model_data)
  colnames(design) <- levels(model_data$ModelGroup)
  contrast <- limma::makeContrasts(
    contrasts = paste0(condition_code[[target]], " - ", condition_code[[reference]]),
    levels = design
  )

  fit <- limma::lmFit(mat, design) |>
    limma::contrasts.fit(contrast) |>
    limma::eBayes()

  result <- data.table::as.data.table(
    limma::topTable(
      fit,
      number = Inf,
      adjust.method = adj_m,
      sort.by = "P"
    ),
    keep.rownames = "ProteinID"
  )

  detection <- data.table::data.table(
    ProteinID = rownames(mat),
    N_detected_reference = n_ref[keep],
    N_detected_target = n_target[keep]
  )

  result <- merge(result, detection, by = "ProteinID", all.x = TRUE)
  result <- merge(result, prot_anno, by = "ProteinID", all.x = TRUE)
  result[, `:=`(
    ComparisonShortID = cmp_i$ComparisonShortID,
    ComparisonID = cmp_i$ComparisonID,
    Target = target,
    Reference = reference,
    Significant = adj.P.Val < alpha & abs(logFC) >= lfc_thr
  )]
  result[, Direction := data.table::fcase(
    Significant & logFC > 0, "Higher_in_target",
    Significant & logFC < 0, "Higher_in_reference",
    default = "Not_significant"
  )]

  result <- result[, .(
    ComparisonShortID,
    ComparisonID,
    Target,
    Reference,
    ProteinID,
    Gene.names,
    Protein.names,
    logFC,
    P.Value,
    adj.P.Val,
    Significant,
    Direction,
    N_detected_reference,
    N_detected_target,
    Peptides_used
  )]

  summary <- data.table::data.table(
    ComparisonShortID = cmp_i$ComparisonShortID,
    ComparisonID = cmp_i$ComparisonID,
    Target = target,
    Reference = reference,
    Exploratory = cmp_i$Exploratory,
    N_reference = length(ref_bio),
    N_target = length(target_bio),
    Proteins_tested = nrow(result),
    Significant_total = sum(result$Significant),
    Higher_in_target = sum(result$Direction == "Higher_in_target"),
    Higher_in_reference = sum(result$Direction == "Higher_in_reference")
  )

  list(result = result, summary = summary)
}
```

# Run comparisons

```{r run_comparisons}
analysis <- lapply(seq_len(nrow(cmp)), function(i) run_comparison(cmp[i]))

res_lst <- stats::setNames(
  lapply(analysis, `[[`, "result"),
  cmp$ComparisonShortID
)

cmp_sum <- data.table::rbindlist(lapply(analysis, `[[`, "summary"))
res_all <- data.table::rbindlist(res_lst)
sig <- res_all[Significant == TRUE][order(ComparisonID, adj.P.Val, -abs(logFC))]
```

# Save results

```{r save_results}
data.table::fwrite(cmp_sum, file.path(pth$da_table, "comparison_summary.csv"))
data.table::fwrite(res_all, file.path(pth$da_table, "all_results.csv"))
data.table::fwrite(sig, file.path(pth$da_table, "significant_results.csv"))

obj <- list(
  comparison_summary = cmp_sum,
  all_results_list = res_lst,
  all_results = res_all,
  sig_results = sig,
  peptide_bio_mat = pep_bio_m,
  protein_bio_mat = p_bio_m,
  bio_info = bio,
  design_all = raw0,
  design_keep = raw,
  prot_annot = prot_anno,
  condition_order = cond,
  settings = list(
    padj_method = adj_m,
    alpha = alpha,
    fc_cutoff = lfc_thr,
    min_detected = min_det,
    batch_variable = "Groups"
  )
)
```
