---
title: "Plotting v9: differential-analysis plots"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
---

# Purpose

This is step 4 of the v9 pipeline. It uses the differential-analysis object created by `3.differential_analysis_v9.Rmd` to generate volcano plots, a significant-protein summary, a heatmap, and protein-level PCA.

# Settings

```{r settings}
load_project_packages(c("data.table", "ggplot2", "matrixStats", "pheatmap"))

cmp_sum <- data.table::copy(obj$comparison_summary)
res_lst <- obj$all_results_list
sig <- data.table::copy(obj$sig_results)
prot_m <- obj$protein_bio_mat
bio <- data.table::copy(obj$bio_info)
cond <- as.character(obj$condition_order)

alpha <- obj$settings$alpha
lfc_thr <- obj$settings$fc_cutoff
plot_font_size <- 16

condition_colors <- stats::setNames(
  c(
    "#0072B2", "#D55E00", "#009E73", "#CC79A7",
    "#E69F00", "#56B4E9", "#F0E442", "#000000"
  )[seq_along(cond)],
  cond
)
```

# Plot functions

```{r plot_functions}
save_plot <- function(plot, name, width, height) {
  ggplot2::ggsave(
    file.path(pth$da_plots, paste0(name, ".png")),
    plot,
    width = width,
    height = height,
    dpi = 300
  )
}

make_pca <- function(mat, sample_info) {
  sample_ids <- sample_info$BioExperiment
  mat <- mat[, sample_ids, drop = FALSE]

  row_median <- matrixStats::rowMedians(mat, na.rm = TRUE)
  keep <- is.finite(row_median)
  mat <- mat[keep, , drop = FALSE]
  row_median <- row_median[keep]

  missing <- which(!is.finite(mat), arr.ind = TRUE)
  mat[missing] <- row_median[missing[, 1]]
  mat <- mat[matrixStats::rowVars(mat) > 0, , drop = FALSE]

  pca <- stats::prcomp(t(mat), center = TRUE, scale. = FALSE)
  variance <- pca$sdev^2 / sum(pca$sdev^2)

  scores <- data.table::as.data.table(
    pca$x[, 1:3, drop = FALSE],
    keep.rownames = "BioExperiment"
  )
  scores <- merge(scores, sample_info, by = "BioExperiment", sort = FALSE)

  attr(scores, "variance") <- variance
  attr(scores, "features_used") <- nrow(mat)
  scores
}
```

# Volcano plots

```{r volcano_plots}
plot_volcano <- function(result) {
  result <- data.table::copy(result)
  result[, neg_log10_adj_p := -log10(pmax(adj.P.Val, .Machine$double.xmin))]
  result[, Call := data.table::fcase(
    Significant & logFC > 0, "Higher in target",
    Significant & logFC < 0, "Higher in reference",
    default = "Not significant"
  )]
  result[, Call := factor(
    Call,
    levels = c("Higher in target", "Higher in reference", "Not significant")
  )]

  comparison <- result$ComparisonID[1]
  short_id <- result$ComparisonShortID[1]

  p <- ggplot2::ggplot(
    result,
    ggplot2::aes(x = logFC, y = neg_log10_adj_p, color = Call)
  ) +
    ggplot2::geom_point(alpha = 0.70, size = 1.4) +
    ggplot2::geom_vline(
      xintercept = c(-lfc_thr, lfc_thr),
      linetype = "dashed",
      linewidth = 0.3
    ) +
    ggplot2::geom_hline(
      yintercept = -log10(alpha),
      linetype = "dashed",
      linewidth = 0.3
    ) +
    ggplot2::scale_color_manual(values = c(
      "Higher in target" = "#D55E00",
      "Higher in reference" = "#0072B2",
      "Not significant" = "#999999"
    )) +
    ggplot2::labs(
      title = paste0("Volcano plot: ", comparison),
      x = "log2 fold change",
      y = "-log10 adjusted p value",
      color = "Call"
    ) +
    ggplot2::theme_bw(base_size = plot_font_size)

  save_plot(p, paste0("volcano_", short_id), width = 12, height = 10)
}

invisible(lapply(res_lst, plot_volcano))
```

# Significant-protein summary

```{r significant_summary}
cmp_sum[, ComparisonLabel := factor(
  paste0(ComparisonShortID, ": ", ComparisonID),
  levels = paste0(ComparisonShortID, ": ", ComparisonID)
)]

p_sig <- ggplot2::ggplot(
  cmp_sum,
  ggplot2::aes(x = ComparisonLabel, y = Significant_total)
) +
  ggplot2::geom_col(width = 0.75, fill = "#0072B2") +
  ggplot2::labs(
    title = "Significant proteins by comparison",
    x = "Comparison",
    y = "Significant proteins"
  ) +
  ggplot2::theme_bw(base_size = plot_font_size) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 60, hjust = 1, vjust = 1)
  )

save_plot(p_sig, "significant_proteins", width = 14, height = 8)
p_sig
```

# Significant-protein heatmap

```{r significant_heatmap}
sig_rank <- sig[, .(
  Best_adjusted_P = min(adj.P.Val, na.rm = TRUE)
), by = ProteinID][order(Best_adjusted_P)]

selected <- intersect(head(sig_rank$ProteinID, 60), rownames(prot_m))
heat_m <- prot_m[selected, , drop = FALSE]
heat_m <- heat_m[rowSums(!is.na(heat_m)) >= 3, , drop = FALSE]

row_median <- matrixStats::rowMedians(heat_m, na.rm = TRUE)
missing <- which(is.na(heat_m), arr.ind = TRUE)
heat_m[missing] <- row_median[missing[, 1]]
heat_m <- t(scale(t(heat_m)))
heat_m[!is.finite(heat_m)] <- 0

heat_annotation <- as.data.frame(
  bio[match(colnames(heat_m), BioExperiment), .(Condition)]
)
rownames(heat_annotation) <- colnames(heat_m)

grDevices::png(
  file.path(pth$da_plots, "significant_protein_heatmap.png"),
  width = 3300,
  height = 3000,
  res = 300
)
pheatmap::pheatmap(
  heat_m,
  annotation_col = heat_annotation,
  annotation_colors = list(Condition = condition_colors),
  show_rownames = FALSE,
  show_colnames = TRUE,
  fontsize = 12,
  fontsize_col = 11,
  main = "Significant proteins across biological experiments"
)
grDevices::dev.off()
```

# Protein-level PCA

```{r protein_pca}
prot_pca <- make_pca(prot_m, bio)
prot_variance <- attr(prot_pca, "variance")

p_prot <- ggplot2::ggplot(
  prot_pca,
  ggplot2::aes(x = PC1, y = PC2, color = Condition)
) +
  ggplot2::geom_point(size = 3.2, alpha = 0.90) +
  ggplot2::scale_color_manual(values = condition_colors, drop = FALSE) +
  ggplot2::labs(
    title = "Protein-level PCA by biological experiment",
    x = paste0("PC1 (", round(100 * prot_variance[1], 1), "%)"),
    y = paste0("PC2 (", round(100 * prot_variance[2], 1), "%)"),
    color = "Condition"
  ) +
  ggplot2::theme_bw(base_size = plot_font_size) +
  ggplot2::theme(legend.position = "right")

save_plot(p_prot, "protein_pca_2d", width = 10, height = 7)
data.table::fwrite(prot_pca, file.path(pth$da_table, "protein_pca_scores.csv"))
p_prot
```

# Three-dimensional protein-level PCA

```{r protein_pca_3d}
load_project_packages("scatterplot3d")

prot_pca[, Condition := factor(as.character(Condition), levels = cond)]
point_colors <- unname(condition_colors[as.character(prot_pca$Condition)])

centroids <- prot_pca[, .(
  PC1 = mean(PC1),
  PC2 = mean(PC2),
  PC3 = mean(PC3)
), by = Condition]

grDevices::png(
  file.path(pth$da_plots, "protein_pca_3d.png"),
  width = 3300,
  height = 2800,
  res = 300
)

graphics::layout(matrix(c(1, 2), nrow = 2), heights = c(4.5, 1))
graphics::par(mar = c(4.5, 4.5, 4, 3))

p_3d <- scatterplot3d::scatterplot3d(
  prot_pca$PC1,
  prot_pca$PC2,
  prot_pca$PC3,
  color = point_colors,
  pch = 16,
  cex.symbols = 1.6,
  angle = 55,
  grid = TRUE,
  box = TRUE,
  cex.axis = 0.9,
  cex.lab = 1.1,
  xlab = paste0("PC1 (", round(100 * prot_variance[1], 1), "%)"),
  ylab = paste0("PC2 (", round(100 * prot_variance[2], 1), "%)"),
  zlab = paste0("PC3 (", round(100 * prot_variance[3], 1), "%)"),
  main = "Protein-level PCA"
)

for (condition_i in as.character(centroids$Condition)) {
  condition_center <- centroids[as.character(Condition) == condition_i]
  condition_color <- unname(condition_colors[condition_i])

  p_3d$points3d(
    condition_center$PC1,
    condition_center$PC2,
    condition_center$PC3,
    pch = 18,
    cex = 2.2,
    col = condition_color
  )
}

graphics::par(mar = c(0, 0, 0, 0))
graphics::plot.new()
graphics::legend(
  "center",
  legend = c(cond, "Condition centroid"),
  col = c(unname(condition_colors[cond]), "black"),
  pch = c(rep(16, length(cond)), 18),
  title = "Condition",
  bty = "n",
  ncol = 3,
  cex = 0.9,
  pt.cex = c(rep(1.2, length(cond)), 1.6)
)

grDevices::dev.off()
```
